Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-11-19 manuscript/manuscript.tex
Top to bottom read through
A number of minor edits, formatting, etc. Nearly done.

\hyphenpenalty=900

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. The City of New York (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, leading to the creation of the NYC Open Data
portal. This portal now hosts 2,700 datasets from 80 city agencies,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. The 
success of these initiatives highlights the importance of data 
curation in ensuring the utility and reliability of open datasets.

@@

In the early 21st century, the open data movement began 
to take shape, driven by the fundamental belief that 
freely accessible data can transform both societies and 
governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly after,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
breaking down barriers that once kept valuable information 
regarding government performance in silos. Their collective impact 
extends beyond mere data sharing to fostering a culture of openness 
that benefits individuals, communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

The City of New York (NYC) has emerged as a leader in the open data movement,
marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2,700 datasets
from 80 different city agencies. This resource has become invaluable
for researchers across various fields and has significantly enhanced
local government transparency. Popular datasets include information on
restaurant health inspection violations, car crashes, high school and
college enrollment statistics, jail inmate charges, and the location
of city\mbox{-}wide free Internet access points. These datasets have been
applied in civil life in various ways, such as mapping car crashes
involving pedestrians and visualizing high school and college
enrollment trends. Furthermore, they have enabled significant research
across multiple domains, including health \citep{cantor2018facets,
  shankar2021data}, urban development \citep{neves2020impacts}, and
transportation \citep{gerte2019understanding}, aiding in the
understanding and addressing of complex urban challenges.

Data curation, the process of organizing, maintaining, and ensuring
the quality of datasets, plays a crucial role in maximizing the
utility of open data. Proper curation ensures that datasets remain
consistent, accurate, and useful for diverse applications. For example, 
well\mbox{-}curated data is essential for machine learning systems, which 
require high\mbox{-}quality data to produce reliable insights 
\citep{polyzotis2019data, jain2020overview}. A critical component of 
data curation is data cleaning, which involves identifying and rectifying 
inconsistencies, errors, and inaccuracies in datasets. While open data 
initiatives have made vast amounts of data available, ensuring the 
reliability of such data hinges on rigorous data cleaning processes. 
Efficient usage of software tools can significantly streamline this 
aspect of curation \citep[e.g.,][]{cody2017cody, van2018statistical}. Poor 
curation may result in issues such as missing data, formatting errors, 
or inconsistent values, leading to biased or inaccurate outcomes 
\citep{geiger2020garbage}. This is particularly critical in domains 
where machine learning is applied to sensitive tasks, such as public 
health or policy \citep{rahm2000data}.

@@

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over a 2\mbox{-}year period from 2022 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
 also investigates field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines some concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}.

@@

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non\mbox{-}emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non\mbox{-}emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone\mbox{-}based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID\mbox{-}19 pandemic, the 2021 expansion to include the MTA's city
subway system (the biggest expansion in their history), and a 
record 3.23 million service requests in 2023.

Today, the NYC 311 data system manages over 3 million service
requests (SRs) per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo\mbox{-}mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, provides technological support for the
open data system and data portal. Input data for the 311 system 
is sourced from 17 different NYC Agencies. Some of these Agencies
utilize the core 311 software, while others transfer data to the
311 Open Data Warehouse from in\mbox{-}house systems.

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping ity governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID\mbox{-}19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies

Our investigation utilizes an SR dataset covering  2\mbox{-}years (2022-2023).
This dataset is approximately 3.7 GB in size and contains over 
6.4 million rows with each row representing one SR. It contains 
41 columns of data and was collected on 15 September 2023. The 
dataset is used to conduct detailed analysis 
in areas such as data consistency, data validity, redundant fields, 
etc.. Instructions on downloading this file from the 
NYC Open Data Portal are available in the supplemental material.
Here are some characteristics of the 311 SR data set:

\begin{itemize}[left=1.5em]
	\item There are 41 columns of data for each row, exportable as a CSV file.
	Each row represents a single Service Request (SR).

	\item There are four date fields 
	(\texttt{created\_date}, \texttt{closed\_date}, \texttt{resolution\_action\_updated\_date}, \& \texttt{due\_date}).

	\item There are two borough fields; one of which appear to be a duplicate.

@@

	\item In addition to  \texttt{incident\_address}, there are five additional location fields: 
	\texttt{latitude \& longitude}, \texttt{street\_name}, \texttt{landmark}, USGS State Plane 
	Coordinate System, and \texttt{block} \#.

	\item A free\mbox{-}form text field, \texttt{resolution\_description}, which 
	supports 934 characters of input, including commas and special characters.
\end{itemize}

During the 10\mbox{-}year period from 2014 to 2023,  311 experienced
a 50\% increase in the number of Service Requests (SRs), 
reflecting a growing public reliance and usage of 
the 311 system. (During this 10\mbox{-}year time-frame, the 
population of NYC grew just 0.6\%.). The most significant increase
in activity occurred in 2020, when the volume of requests 
surged to 3.23 million during the COVID\mbox{-}19 pandemic, 
illustrating the essential role of the 311 system in supporting 
NYC residents during times of crisis. This steady 
rise in SRs can be at least partially attributed to the increased 
accessibility of the 311 system via online and mobile 
platforms, as well as heightened public awareness of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in SR volume in future years,  
highlighting the system's expanding role in managing 
both routine City operations and extraordinary events.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
the responsible agency, showing the cumulative 
percentage of SRs handled by each
agency over a the 2\mbox{-}year time-frame. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

\begin{itemize}[left=1.5em]
    \item New York Police Department (NYPD): 43\% of all SRs
    \item Housing Preservation and Development (HPD): 21\%
    \item New York City Department of Sanitation (DSNY): 10\%
    \item Department of Transportation (DOT): 7\%
    \item Department of Environmental Protection (DEP): 5\%
    \item Department of Parks and Recreation (DPR): 4\%
\end{itemize}

Collectively these six City Agencies handle over 90\% of the total SR 
volume, indicative of the critical role these Agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The remaining
10\% of SRs is distributed across 10 additional Agencies. 
This concentration of SRs to the ``big six'' Agencies 
underscores the necessity for optimal data handling and 
curation processes within these six high volume Agencies.

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

Figure~\ref{fig:SR_complaints}
illustrates how the top 20 complaint types account for 70\% 
of all SRs. Note that the distribution of complaints is skewed, 
with a small number of issues dominating total SR volume. 
Of special interest are noise\mbox{-}related complaints, of which there are eight different
types including residential, commercial, and 
street noise. When combined noise\mbox{-}related SRs make up 22\% of all 
complaints; the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
reveals that the top 20 complaint types comprise 70\% of SRs,
while the remaining 190 complaint types are spread thinly 
across the remaining 30\%. This suggests that improving 
responses to the most frequent complaints could have an 
outsized impact on overall service efficiency and resident 
satisfaction. This chart also provides insights into the 
operational pressures faced by City Agencies 
responsible for handling these high-volume complaints.

@@

Many quality criteria are employed to ensure high\mbox{-}quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats; typically 
outlined in a \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}. An 
essential aspect of validation is domain compliance, where 
fields must adhere to a predefined set of 
values such as statuses, state names, or zip codes. Structural 
errors can also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary,  
inconsistent use of data fields. Furthermore, redundant 
or irrelevant fields can clutter datasets, reducing efficiency, 
creating confusion, and introducing errors. Logical inconsistencies 
are another consideration, such as related fields that violate expected 
relationships, e.g. a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure meaningful data.

We identify the presence of issues in the NYC 311 SR open dataset;
not attempting to solve them, but rather to highlight their scope, 
magnitude, and potential impact. We would recommend 
that a corrective solution should be undertaken only 
after further investigation as to the why and how such 
issues might come about, including  detailed discussions 
with the originating agency as to whether or not it is an 
actual error, correct,  or holds some other status.

@@

\paragraph{Undocumented Fields:} The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' and include:

@@

While one can likely infer the meaning of these ``computed'' 
fields, uncertainty remains as to their derivation and 
usage. These six fields are not included in
the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}. Additionally, 
our analysis shows that these six ``computed'' fields have 
significant data validity issues, leading us to treat these six fields 
as experimental and \textit{not for suitable for public use}. An update 
to the Data Dictionary would seem appropriate. Accordingly, 
we will not address these fields in our analysis.

\paragraph{Representation of Missing data:} We found that the 
311 SR dataset exhibited inconsistent approaches when 
presenting missing data. We discovered many different 
representations of missing data: nulls, spaces, ``NA'', ``N/A'', 
and ``<NA>''. This variety of representations complicates programming
efforts, and increases the likelihood of introducing errors 
in data analysis (typically by search for or excluding missing fields 
and not aware of the various representations). A more consistent 
approach to treating missing data should be employed to address this issue.

\paragraph{Challenges with daylight saving time (DST):} The 311 SR 
dataset captures local NYC times. Accordingly, twice each 
year there are changes to the local time as a result of observing 
DST, and this can introduce some unexpected results. For example, 
on March 13, 2022 and March 12, 2023 there are no times 
between 1:59AM and 2:59AM: the 2:00AM hour on those dates 
is simply missing owing to the clocks moving forward; at 2:00AM 
it instantly becomes 3:00AM. This creates a gap in the hourly 
distribution of SRs on those days, and it exaggerates the 
lifespan of SRs, albeit by a single hour. Similarly on 
November 6, 2022 and November 5, 2023, the 
hour of 1:00AM - 2:00AM is repeated twice. This can result 
in a bizarre scenario where, for example, an SR can 
be created at 1:59AM and then closed say 30 minutes later 
at 1:29AM; seemingly nonsensical. Utilizing UTC time 
for date\mbox{-}time stamps might help address this issue.

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} field are blank. When sorting fields by
blank (or N/A values) they appear to divide into three groups:

@@

The \textit{Mostly Empty} category includes fields such as
\texttt{taxi\_company\_borough}, \texttt{due\_date},
\texttt{road\_ramp}, and \texttt{bridge\_highway\_name}.
The \textit{Partially Empty} category includes fields such as
\texttt{location\_type}, \texttt{landmark}, 
and \texttt{cross\_street\_1 \& \_2}. And the \textit{Few/None Empty} 
category includes \texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. For some analysis efforts, it may be
prudent to inquire further as to why some fields are 
mostly or almost always blank. Figure~\ref{fig:blank_fields} is 
a graphic depiction of total empty (blank \& N/As) for each field illustrating
the grouping into the Most, Partial, and Few categories.

@@

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. The \texttt{latitude} 
and \texttt{longitude} fields were found to fall within the 
geographic boundaries of New York City, and 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} specifies very few domains of 
acceptable values for other fields. Nonetheless, the following fields were tested and found to comply with their expected domains, as determined by public 
usage and as evidenced in other historical datasets:

@@

\paragraph{Zip Codes:}
\label{sec:zipcodesissues}
All \texttt{incident\_zip} zip codes should 
validate against the USPS database, which contains 
44,173 valid zip codes. And while the \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by agency, it mirrors the \textit{overall} 
breakdown of SRs by agency, potentially 
indicating a systemic problem.

\paragraph{Created and Closed Dates}
\label{sec:negativeduration}
With the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields would be automatically populated by the  
application software. Thus when updating an SR status to ``new'' or ``closed''
the appropriate software should automatically populate those fields. 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the \texttt{created\_date} and \texttt{closed\_date} fields that
an automated process would normally exclude, including:

>>>>

\hyphenpenalty=950

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. The City of New York (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, leading to the creation of the NYC Open Data
portal. This portal currently hosts 2,700 datasets,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. The 
success of this initiative highlights the importance of data 
curation in ensuring the utility and reliability of open datasets.

@@

In the early 21st century, the open data movement began 
to take shape, driven by the fundamental belief that 
freely accessible data can transform both societies and 
governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly afterwards,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
removing barriers that once kept valuable information 
regarding government performance in silos. Their collective impact 
extends beyond mere data sharing to fostering a culture of openness 
that benefits individuals, communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

The City of New York (NYC) has emerged as a leader in the open data movement,
marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2,700 datasets
across 80 different city agencies. This resource has become invaluable
for researchers across various fields and has significantly enhanced
local government transparency. Popular datasets include information on
restaurant health inspection violations, car crashes, high school and
college enrollment statistics, jail inmate charges, and the location
of City\mbox{-}wide free Internet access points. These datasets have been
applied in civil life in various ways, such as mapping car crashes
involving pedestrians and visualizing high school and college
enrollment trends. Furthermore, they have enabled significant research
across multiple domains, including health \citep{cantor2018facets, 
shankar2021data}, urban development \citep{neves2020impacts}, and
transportation \citep{gerte2019understanding}, aiding in the
understanding and addressing of complex urban challenges.

Data curation, the process of organizing, maintaining, and ensuring
the quality of datasets, plays a crucial role in maximizing the
utility of open data. Proper curation ensures that datasets remain
consistent, accurate, and useful for diverse applications. For example, 
well\mbox{-}curated data is essential for machine learning systems, which 
require high\mbox{-}quality data to produce reliable insights 
\citep{polyzotis2019data, jain2020overview}. A critical component of 
data curation is data cleaning, which involves identifying and rectifying 
inconsistencies, errors, and inaccuracies in datasets. While open data 
initiatives have made vast amounts of data available, ensuring the 
reliability and utility of such data hinges on rigorous data cleaning processes. 
Efficient usage of software tools can significantly streamline this 
aspect of curation \citep[e.g.,][]{cody2017cody, van2018statistical}. Poor 
curation may result in issues such as missing data, formatting errors, 
or inconsistent values, leading to biased or inaccurate outcomes 
\citep{geiger2020garbage}. This is particularly critical in domains 
where machine learning is applied to sensitive tasks, such as public 
health or policy \citep{rahm2000data}.

@@

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over a 2\mbox{-}year period from 2022 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
 also investigates data field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines some concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}.

@@

\section{NYC 311 Service Request (SR) Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non\mbox{-}emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non\mbox{-}emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone\mbox{-}based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 associated 
with the the COVID\mbox{-}19 pandemic, the 2021 expansion 
to include the MTA's city subway system (the biggest 
expansion in 311  history), and a record 3.23 million 
service requests in 2023.

Today, the NYC 311 data system manages over 3 million service
requests (SRs) per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo\mbox{-}mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology 
and Innovation (OTI)}, provides technical support for 
the 311 application, the open data infrastructure, mobile \& web applications, 
and the public\mbox{-}facing data portal. Input data for the 311 system 
is sourced from 16 different NYC Agencies. Some of these Agencies
utilize the core 311 software, while others transfer data to the
311 Open Data Warehouse from their in\mbox{-}house systems.

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping City governance and community
engagement. Open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID\mbox{-}19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies.

Our investigation utilizes a 311 SR dataset covering  2\mbox{-}years (2022-2023).
 This sample was collected on 15 September 2024, and is used 
 to conduct detailed analysis in areas such as data 
 consistency, data validity, redundant fields, etc.. Instructions 
 on downloading this file from the NYC Open Data Portal 
 are available in the supplemental material. Here are some 
 characteristics of the 311 SR dataset used in this analysis:

\begin{itemize}[left=1.5em]
	\item The dataset is approximately 3.7 GB in size and contains 
	over 6.4 million rows. It was exported as a CSV file.

	\item There are 41 columns of data for each row, with each
	row representing a single Service Request (SR).

	\item There are four date fields: \texttt{created\_date}, \texttt{closed\_date}, 
	\texttt{resolution\_action\_updated\_date}, \& \texttt{due\_date}.

	\item There are two borough fields; one of which appears to be a duplicate.

@@

	\item In addition to  \texttt{incident\_address}, there are five additional location fields:

	\texttt{latitude \& longitude}, \texttt{street\_name}, \texttt{landmark}, USGS State Plane 
	Coordinate System, and \texttt{block} \textbf{\#}.

	\item A free\mbox{-}form text field, \texttt{resolution\_description}, which 
	supports 930 characters of input, including commas and 
	special characters  which can be problematic for automated processing.
\end{itemize}

During the 10\mbox{-}year period from 2014 to 2023,  NYC's 311 system
experienced a 50\% increase in the number of SRs, 
reflecting a growing public reliance on and usage of 
the 311 system. (During this 10\mbox{-}year time-frame, the 
population of NYC grew just 0.6\%.). The most significant increase
in activity occurred in 2020, when the volume of requests 
surged to 3.23 million during the COVID\mbox{-}19 pandemic, 
illustrating the essential role of the 311 system in supporting 
NYC residents during times of crisis. This steady 
rise in SRs can be at least partially attributed to the increased 
accessibility of the 311 system via online and mobile 
platforms, as well as heightened public awareness of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in SR volume in future years,  
highlighting the system's expanding role in managing 
both routine City operations and extraordinary events.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
the responsible Agency, showing the cumulative 
percentage of SRs handled by each
agency over the 2\mbox{-}year time-frame. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

\begin{enumerate}[left=1.5em]
    \item New York Police Department (NYPD): 43\% of all SRs
    \item Housing Preservation and Development (HPD): 21\%
    \item New York City Department of Sanitation (DSNY): 10\%
    \item Department of Transportation (DOT): 7\%
    \item Department of Environmental Protection (DEP): 5\%
    \item Department of Parks and Recreation (DPR): 4\%
\end{enumerate}

Collectively these six City Agencies handle over 90\% of the total SR 
volume, indicative of the critical role these Agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The remaining
10\% of SRs is distributed across 10 additional Agencies. 
This concentration of SRs to the ``big six'' Agencies 
underscores the necessity for optimal data handling and 
curation processes within these  high volume Agencies.

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{SRs_by_Agency.pdf}
  	\caption{SR Counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

The dataset contains 210 different \texttt{complaint\_types}. 
Figure~\ref{fig:SR_complaints} illustrates how the Top 20 
\texttt{complaint\_type(s)} account for 70\% 
of all SRs. The distribution of complaints is skewed, 
with a small number of issues dominating total SR volume. 
Of special interest are noise\mbox{-}related complaints, of 
which there are eight different types including vehicle, 
residential, commercial, and street noise. When 
combined, noise\mbox{-}related SRs make up 22\% of all 
complaints; the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
reveals that the Top 20 \texttt{complaint\_type(s)} comprise 70\% of SRs,
while the remaining 190 other \texttt{complaint\_type(s)}  are spread thinly 
across the remaining 30\%. This suggests that improving 
responses to the most frequent complaints could have an 
outsized impact on overall service efficiency and resident 
satisfaction. This chart also provides insights into the 
operational pressures faced by those City Agencies 
responsible for handling these high-volume complaints.

@@

Many quality criteria are employed to ensure high\mbox{-}quality data 
processing. One of the primary efforts is data validation, which includes 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats; typically 
outlined in a \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}. An 
essential aspect of validation is domain compliance, where 
fields must adhere to a predefined set of 
values such as statuses, state names, or zip codes. Structural 
errors can also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary and  
inconsistent usage of data fields. Furthermore, redundant 
(or irrelevant) fields can clutter datasets, reducing efficiency, 
creating confusion, and introducing errors. Logical inconsistencies 
are another consideration, such as related fields that violate expected 
relationships, e.g. a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be managed to ensure meaningful data.

We identify the presence of issues in the NYC 311 SR open dataset;
not attempting to solve them, but rather to highlight their scope, 
magnitude, and potential impact. We would recommend 
that a corrective solution should be undertaken only 
after further investigation as to the why and how such 
issues might come about, including  detailed discussions 
with the originating agency as to whether or not the issue is an 
actual error, correct,  or holds some other status.

@@

\paragraph{Undocumented Fields:} The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded if manually selected. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' and include:

@@

While one can likely infer the meaning of these ``computed'' 
fields, uncertainty remains as to their derivation and 
usage. These six fields are not included in
the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}. Additionally, 
our analysis shows that these six ``computed'' fields have 
significant data validity issues, leading us to categorize these six fields 
as experimental and \textit{not for public use}. An update 
to the Data Dictionary would seem appropriate. Accordingly, 
we will not address these fields in our analysis.

\paragraph{Representation of Missing data:} We found that the 
311 SR dataset exhibited inconsistent approaches when 
presenting missing data. We discovered many different 
representations of missing data: nulls, spaces, ``NA'', ``N/A'', 
and ``<NA>''. This variety of representations complicates programming
efforts, and increases the likelihood of introducing errors 
in data analysis (typically by search for or excluding missing fields 
and not aware of the various representations of such). A more consistent 
approach to treating missing data should be employed to address this issue.

\paragraph{Challenges with daylight saving time (DST):} The 311 SR 
dataset captures local NYC times. Accordingly, twice each 
year there are changes to the local time as a result of observing 
DST. This can introduce some unexpected results. For example, 
on both March 13, 2022 and March 12, 2023 there are no times 
between 1:59AM and 2:59AM. The 2:00AM hour on those dates 
is simply missing owing to the clocks moving forward; at 2:00AM 
it instantly becomes 3:00AM. This creates a gap in the hourly 
distribution of SRs on those days, and exaggerates the 
lifespan of SRs, albeit by a single hour. Similarly on both
November 6, 2022 and November 5, 2023, the 
hour of 1:00AM - 2:00AM is repeated twice. This can result 
in a bizarre scenario where, for example, an SR can 
be \textit{created} at 1:59AM and then be \textit{closed} 30 minutes later 
at 1:29AM; seemingly nonsensical. Utilizing UTC time 
for date\mbox{-}time stamps would help address this issue.

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} fields are blank. When sorting fields by
blank (or N/A values) they appear to divide into three groups:

@@

The \textit{Mostly Empty} category includes fields such as
\texttt{taxi\_company\_borough}, \texttt{due\_date},
\texttt{road\_ramp}, and \texttt{bridge\_highway\_name}.
The \textit{Partially Empty} category includes fields such as
\texttt{location\_type}, \texttt{landmark}, 
and \texttt{cross\_street\_1 \& \_2}. And the \textit{Few/None Empty} 
category includes \texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. For some analysis efforts, it may be
prudent to inquire further as to why some fields are 
mostly or almost always blank. Figure~\ref{fig:blank_fields} is 
a graphic depiction of total empty (blank \& N/As) for each 
field illustrating the grouping into 
the \textbf{Most}, \textbf{Partial}, and \textbf{Few/None} categories.

@@

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. For example, the \texttt{latitude} 
and \texttt{longitude} fields were found to all fall within the 
geographic boundaries of New York City. Likewise, 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} specifies very few domains of 
acceptable values for other fields. Nonetheless, the following fields were tested and found to comply with their expected domains, as determined by public 
usage and as evidenced in other historical datasets:

@@

\paragraph{Zip Codes:}
\label{sec:zipcodesissues}
All \texttt{incident\_zip} zip codes should 
validate against the USPS database, which contains 
44,173 valid zip codes. We discovered that the \texttt{incident\_zip} 
field has 0.07\% invalid entries, resulting in 4,374 errors. This includes
zip codes such as ``12345'', ``10000'', and ``99999'. If you group 
the invalid zip code entries by percentage by agency, it 
mirrors the \textit{overall} breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\paragraph{Created and Closed Dates:}
\label{sec:negativeduration}
With the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields would be automatically populated by the  
application software. Thus when creating  or closing an SR 
the the software would automatically populate these fields. 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the \texttt{created\_date} and \texttt{closed\_date} fields that
an automated process would normally exclude, including:

\begin{itemize}[left=1.5em]
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} resulting a nonsensical ``negative duration''. (12,450)
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. ``1900-01-01''. (8)
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match \textbf{to the second} creating an impossible ``zero duration''. (163,720)
    \item An unusually large number of SRs closed and/or created 
    exactly at midnight, to the second. (247,000)
\end{itemize}
